---
title: "Stock Report"
author: "Brad Boehmke"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
params:
  symbol:
    choices:
    - AMZN
    - GOOG
    - FB
    - KR
    - TSLA
    value: AMZN
---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(xts)
library(quantmod)
library(magrittr)

prices <- getSymbols(params$symbol, auto.assign = FALSE)
move <- Cl(last(prices)) - Op(last(prices))

decision <- ifelse(move > 0, "BUY", "SELL")
```

## `r params$symbol` Recommendation: `r decision`

Based on our internal trading algorithm, we believe that the **`r params$symbol`** stock will **`r ifelse(move > 0, "increase", "decrease")`** in price during the next trading period. 


## Price History

The chart below is made with the `quantmod` R package, a widely used package for collecting and visualizing financial data in R. You can learn more about `quantmod` at the website [www.quantmod.com](www.quantmod.com).  

```{r, echo=FALSE, fig.align='center', fig.height=4, fig.width=6}
chartSeries(prices, theme = chartTheme("white", bg.col = "white"))
```

## Method

This forecast was predicted with the *recency algorithm*, a simple---probably useless---method for determining stock prices. The recency algorithm predicts that the next price movement, $M_{j}$, will be in the same direction as the most recent price movement. $M_{i}$.

$$M_{i}= Close_{i} - Open_{i} \tag{1}$$

$$
    M_{j}= 
\begin{cases}
    > 0,              & \text{if } M_{i} > 0\\
    \leq 0,              & \text{if } M_{i}\leq 0
\end{cases}
$$

## Raw Data

The table below displays the most recent 10 days of price data for the `r params$symbol` stock.  

```{r, echo=FALSE}
as.data.frame(prices) %>%
  tail(10) %>%
  knitr::kable(caption = "Stock information for previous 10 business days")
```

