---
title: "Writing Functions"
output: html_notebook
---

A function centralizes a common task to a single, abstract method

This helps to:

- increase reusability which...
- reduces verbosity
- reduces repetitiveness
- reduces the chance of making an error
- reduces instances requiring updates when changes are needed
- reduces amount of code to test

# Key ingredients to a function

1. Name
2. arguments
3. body
4. environment

```{r slide-7}
# present value function
pv <- function(fv, r, n = 5) {
  fv / (1 + r)^n
}

# arguments
args(pv)

# body
body(pv)

# environment
environment(pv)
```

## Your Turn!

Identify the arguments, body, and environment of:

1. `read.csv`
2. `dplyr::add_count`
3. `sum`


# Define a function

- We define a function with <- just like we define any other R object
- Use informative names; strive to use verbs when possible
- We can define a function with no arguments; however, this is rarely useful
- Consequently, most functions have all key ingredients

```{r slide-13}
compute_pv <- function(fv, r, n) {
  fv / (1 + r)^n
}

compute_pv(fv = 1000, r = .05, n = 10)

compute_pv(fv = 2000, r = .08, n = 5)
```

## Your Turn!

- Define a function titled ratio that takes arguments x and y and returns their 
  ratio, `x / y`
- Call `ratio()` with arguments 3 and 4

```{r slide-14}
ratio <- function(___, ___) {
  _______
}
```


Calling arguments in different ways

Many ways to call arguments:

- Using argument names
- Positional matching
- Must use names if you change order
- ...otherwise error or incorrect computation will occur
- missing arguments results in error

```{r slide-18}
# using arg names
compute_pv(fv = 1000, r = .05, n = 10)

# positional matching
compute_pv(1000, .05, 10)

# use names when changing order
compute_pv(r = .05, fv = 1000, n = 10)

# ...otherwise an error occurs
compute_pv(.05, 1000, 10)

# missing arguments result in error
compute_pv(1000, .05)
```

# Setting good defaults

- As our functions require more inputs...
- many of them can likely do with defaults
- Usually, the first 1-3 arguments are not preset but the rest are
- Good defaults are largely based on
   - methodology best practices (i.e. hyperparameters, removing missing values)
   - industry best practices or current values (i.e. current interest rates)
   - organizational best practices (i.e. golden rules)

```{r slide-20}
# set default value for digits argument
compute_pv <- function(fv, r, n, digits = 2) {
  round(fv / (1 + r)^n, digits = digits)
}

# by default, will round to 2 digits
compute_pv(1000, .05, 10)

# to adjust, specify by position
compute_pv(1000, .05, 10, 0)

# or by name
compute_pv(1000, .05, 10, digits = 0)
```

## Your Turn Part 1

- Earlier in these slides you saw the following code duplicated:

```
(df$a - min(df$a, na.rm = TRUE)) / (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
```

- Can you write a function called rescale() that takes argument x and executes 
  this code?
- Test it on the following vector:

```{r slide-21}
set.seed(123)
test_vector <- runif(20, min = 25, max = 40)
```

## Your Turn Part 2

- Now add an argument to rescale() that allows you to round the output to a 
  specified decimal.
- Set the default to 2.


# Checking arguments and other conditions

- We've created a function that allows users to specify their inputs
- But how do we ensure they provide us with the right kind of inputs?
- Or what if we want to be able to provide some sort of feedback to the user?
- There are several ways to signal conditions to function users:
   - `stop()` & `stopifnot()`: signal an error (no way for a function to continue 
      and execution must stop)
   - `warning()`: signal a warning (something has gone wrong but the function has 
      been able to at least partially recover.)
   - `message()`: signal an informative message (function works fine but user 
      should be informed of something.)

## `stop()` & `stopifnot()`

- The stop functions are most commonly used to check for proper inputs
- but can be used to stop the function procedures because the user's environment 
  is not properly established
- Weigh the benefits of highly custom, informative error messages (via `stop`) 
  versus short to the point (via 
  `stopifnot`)
- Execution will stop at the first violation

```{r slide-24}
# stop
compute_pv <- function(fv, r, n, digits = 2) {
  if(!is.numeric(fv)) {
    stop("`fv` must be numeric", call. = FALSE)
  }
  if(!is.numeric(r)) {
    stop("`r` must be numeric", call. = FALSE)
  }
  if(!is.numeric(n)) {
    stop("`n` must be numeric", call. = FALSE)
  }
  if(!is.numeric(digits)) {
    stop("`digits` must be numeric", call. = FALSE)
  }
  round(fv / (1 + r)^n, digits = digits)
}

compute_pv("1000", ".05", 10)

# stopifnot
compute_pv <- function(fv, r, n, digits = 2) {
  stopifnot(is.numeric(fv), is.numeric(r), is.numeric(n), is.numeric(digits))
  round(fv / (1 + r)^n, digits = digits)
}
compute_pv("1000", ".05", 10)
```

## `warning()`

- `warnings()` are rarely used but...
- can be useful to signal non-terminating concerns
   - hyperparameter clashes in a grid search
   - an extremely long compute time and the user may want to re-assess
   - deprecated functions
   - NAs, NULLs, Infs exist, which result in NA output

```{r slide-25}
# example in established function
lubridate::mdy(c("2-1-2019", "1995"))

# how to implement
present_value <- function(fv, r, n, digits = 2) {
  
  warning("`present_value()` is deprecated. Use `compute_pv()`", call. = FALSE)
  stopifnot(is.numeric(fv), is.numeric(r), is.numeric(n), is.numeric(digits))
  
  round(fv / (1 + r)^n, digits = digits)
}

present_value(1000, .05, 10)
```

## `message()`

- designed to be informational
- use them to tell the user that you’ve done something on their behalf
- may represent:
  - default argument that may be non-trivial
  - how information was interpreted by the function
  - words of encouragement!
  - or even industry/business best practices
- Always provide an option to suppress messages

```{r slide-31}
compute_pv <- function(fv, r, n, digits = 2, quiet = FALSE) {
  
  stopifnot(is.numeric(fv), is.numeric(r), is.numeric(n), is.numeric(digits))
  
  if(!quiet) {
    if(r < 0.01 || r > 0.25) { 
    message("Your interest rate exceeds the normal ranges of 1-25%") 
    }
    # other messages here...
  }
  
  round(fv / (1 + r)^n, digits = digits)
}
# default provides message
compute_pv(1000, .26, 10)

# user can silence
compute_pv(1000, .26, 10, quiet = TRUE)
```

## Your Turn!

Let's go back to our `rescale()` function:

```{r slide-32a}
rescale <- function(x, digits = 2){
  rng <- range(x, na.rm = TRUE)
  scaled <- (x - rng[1]) / (rng[2] - rng[1])
  round(scaled, digits = digits)
}
```

- Move the `na.rm = TRUE` option to the function arguments so that the user can 
  control whether or not to remove missing values.
- Add a `warning()` that tells the user NAs are present (if they use `na.rm = FALSE`)
- Add a `message()` that tells the user how many NAs were removed
- add a missing value to your test_vector and test out your revised function

# Lazy evaluation

In R, function arguments are lazily evaluated: only evaluated if accessed

```{r slide-35}
h01 <- function(x) {
  10
}

h01(x = stop("This is an error!"))
```

This provides you with lots of flexibility in when and how to evaluate inputs

```{r slide-36}
h02 <- function(x, y, z) {
  if(missing(z)) {
    sum(x, y)
  } else {
    sum(x, y, z)
  }
}

# z does not exist
h02(x = 1, y = 2)

# z does exist
h02(x = 1, y = 2, z = 3)
```

You need to be careful with how you use this, as it even allows you to use 
arguments outside of your function environment

```{r slide-37}
h03 <- function(x, y) {
    if(exists("z")) {
      x + y + z
    } else {
      x + y
    }
}
# z does not exist
h03(x = 1, y = 2)

# z exists in global environment
z <- 3
h03(x = 1, y = 2)
```

One of the more common applications is to use NULL to indicate that a variable 
is not required but can be used if supplied

```{r slide-38}
compute_pv <- function(fv, r, n, multiplier = NULL) {
  # compute present value
  pv <- round(fv / (1 + r)^n)
  
  # apply multiplier if its supplied
  if(is.null(multiplier)) {
    pv
  } else {
    pv * multiplier
  }
}

# w/o multiplier
compute_pv(1000, .05, 10)

# with multiplier
compute_pv(1000, .05, 10, multiplier = 1.25)
```

# Lexical scoping

- R uses a nested environment structure
- a function has:
   - its own environment
   - plus one or more parent environments
   - always a fresh start
- a function will search:
   - itself
   - through hierarchical parental environments 
- a function will never:
   - use a parameter from a parent environment when it exists in the current 
    environment
   - search within another function unless it's its parent

# Pure Functions

- This is why `h03()` in the previous example worked
- However, we now have an impure function
- Pure function:
   - returns the same result if given the same arguments
   - does not cause any observable side effects

```{r slide-41}
#---------------------BAD---------------------------------------------------
# h03() changes depending on z's value
# outside of the function call
z <- 4
h03(x = 1, y = 2)

z <- 5
h03(x = 1, y = 2)

#---------------------GOOD--------------------------------------------------
# if your function is dependent on randomness
# allow users to set seed
get_n_samples <- function(x, n, seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }
  sample(x, size = n)
}

x <- 1:10

# allows user to always get same results
get_n_samples(x, 5, seed = 123)
get_n_samples(x, 5, seed = 123)
```

```{r slide-42}
# NEVER DO THIS!
get_n_samples <- function(x, n, seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }
  samples <- sample(x, size = n)
  
  # overwrite x
  x <<- samples
  
  # change options to print 8 decimals
  options(digits = 2)
  samples
}

x <- seq(.0001, 1, by = 0.001)
get_n_samples(x, 5, seed = 123)

x
```

Take-away:

- functions should do one thing
- functions should do that one thing consistently
- functions should not do anything else

# Distributing your functions

Two main ways to distribute your functions:

1. using the source() function to read an R script
   - easy to get started
   - all functions/objects in the sourced .R script will be imported to your 
     global environment
   - does not scale well
   
```{r slide-44}
source("my-functions/custom-stat-functions.R")
ls()

purrr::map_dbl(mtcars, skewness)
```

2. writing a package (package writing will not be covered)
   - once you have a few functions that serve a common use
   - hard to get started
   - scales beautifully 
   
# Bells & whistles

We can do many other things with functions such as:

## add progress bars

```{r slide-46}
import_monthly_data <- function(files, quiet = FALSE) {
  if(!quiet) {
    pb <- progress_estimated(length(files))
  }
  special_import <- function(x) {
    if (quiet) {
      suppressMessages(readr::read_csv(x))
    } else {
      pb$tick()$print()
      suppressMessages(readr::read_csv(x))
    }
  }
  purrr::map_dfr(paste0("data/", files), special_import)
}
import_monthly_data(list.files("data"))
```

## take any number of additional arguments with ... to:

- pass arguments on to internal arguments
   
```{r slide-47}
plot_strongest_corr <- function(data, response, ...) {
  # find strong correlation
  vars <- setdiff(names(data), response)
  correlations <- sapply(data[, vars], function(x) cor(data[, response], x))
  x <- names(which.max(abs(correlations)))
  
  # plot relationship
  plot(data[, x], data[, response], xlab = x, ylab = response, ...)
}

plot_strongest_corr(mtcars, "mpg", pch = 19, col = "blue", main = "Some cool title")
```

- or accept unknown number of inputs

```{r slide-48}
select <- function(.data, ...) {
  # convert ... to a list and then to a string
  arguments <- substitute(...())
  vars <- paste(arguments)
  
  # index for 
  .data[, vars]
}

select(mtcars, mpg, wt)
```

- improve efficiency with C++

```{r slide-49}
# create our own sum function
sumR <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total
}

# use Rcpp to write a sum function in C++
library(Rcpp)
cppFunction('double sumC(NumericVector x) {
  int n = x.size();
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}')

x <- runif(1e3)
bench::mark(
  sum(x),
  sumC(x),
  sumR(x)
)[1:6]
```

# Your Turn - Time Permitting

Practice writing the following functions and test them on the given test vector

```{r slide-51}
# test vector
set.seed(123)
x <- rlnorm(100)

# create a function that computes the variance of x
(1 / (length(x) - 1)) * sum((x - mean(x))^2)

# create a function that computes the standard deviation of x
sqrt((1 / (length(x) - 1)) * sum((x - mean(x))^2))

# create a function that computes the standard error of x
var_x <- (1 / (length(x) - 1)) * sum((x - mean(x))^2)
sqrt(var_x / length(x))

# create a function that computes the skewness of x
n <- length(x)
v <- var(x)
m <- mean(x)
third.moment <- (1 / (n - 2)) * sum((x - m)^3)
third.moment / (var(x)^(3 / 2))
```

# Learning More

Advanced R: https://adv-r.hadley.nz/
