---
title: "Iterating with Loops"
output: html_notebook
---

# Intro

Often, we need to execute repetitive code statements a particular number of times. 
Or, we may even need to execute code for an undetermined number of times until a 
certain condition no longer holds.

* `for` loops: execute code for a prescribed number of times (Rule of 👍, 3x).
* `while` & `repeat` loops: repeat code while condition holds
* `next` & `break`: additional clauses to control flow

# `for` loops

Most basic for loop you will see

```{r slide-5}
for (i in 2010:2017) {
  print(paste("The year is", i))
}
```

Re-write the above with `seq_along()` on a pre-defined variable

```{r slide-6}
x <- 2010:2017

for (i in seq_along(x)) {
  print(paste("The year is", x[i]))
}
```

## Your Turn!

We can see all data sets that we have in the "data" folder with `list.files()`

```{r slide-9a}
list.files("data")
```

Say we wanted to import one of these files into R with the following steps

```{r slide-9b}
# here's a single file
(first_df <- list.files("data")[1])

# create path and import this single file
df <- readr::read_csv(paste0("data/", first_df))

# create a new name for file
(new_name <- stringr::str_sub(first_df, end = -5))

# dynamically rename file
assign(new_name, df)
```

Can you incorporate these procedures into a for loop to import all the data 
files?

```{r slide-9c}
# fill in these gaps
for(i in _____) {
  # 1: import data
  df <- readr::read_csv(_____)
  
  # 2: remove ".csv" of name
  new_name <- _____
  
  # 3: dynamically rename file
  assign(_____, _____)
}
```

## Controlling sequences

There are two ways to control the progression of a loop:

1. `next`: terminates the current iteration and advances to the next.
2. `break`: exits the entire for loop.

```{r slide-12}
x <- c(1, 2, NA, 3)

# use next to skip an iteration
for (i in x) {
  if (is.na(i)) next
  print(i)
}

# use break to stop
for (i in x) {
  if (is.na(i)) break 
  print(i)
}
```

Note that break and next only control the progression of the direct loop 
environment they are in.

Break out of inner loop only:

```{r slide-13a}
x <- 1:3
y <- c("a", "b", NA)

for(i in x) {
  print(i)
  for(j in y) {
    if(is.na(j)) break # break out of inner loop
    print(j)
  }
}
```

Use `stop` to stop entire process. You'll learn more about stop() in the functions
module!

```{r slide-13b}
for(i in x) {
  print(i)
  for(j in y) {
    if(is.na(j)) stop("break in inner loop")
    print(j)
  }
}
```

## Your Turn!

The following code identifies the month of the data set:

```{r slide-14a}
# data files
(data_files <- list.files("data"))

# extract month number
as.numeric(stringr::str_extract(data_files, "\\d+"))
```

Modify the following `for` loop with a `next` or `break` statement to:

1. only import Month-01 through Month-07
2. only import Month-08 through Month-10

```{r slide-14b}
# Modify this code chunk with you next/break statement
for(i in list.files("data")) {
  # steps to import each data set
  df <- readr::read_csv(paste0("data/", i))
  new_name <- stringr::str_sub(i, end = -5)
  assign(new_name, df)
  rm(df)
}
```


# Repeating code for undefined number of iterations

Sometimes we need to execute code for an undetermined number of times until a 
certain condition no longer holds.

Two very similar options:

1. `while` loop
   - Test condition first
   - Then execute code
2. `repeat` loop
   - Execute code first
   - Then test condition


```{r slide-17a}
# while loop
coin <- c("heads", "tails")
n_tries <- 0
flip <- NULL

while(length(unique(flip)) != 1) {
  # flip coin 10x
  flip <- sample(coin, 10, replace = TRUE)
  
  # add to the number of tries
  n_tries <- n_tries + 1
}

n_tries
```

```{r slide-17b}
coin <- c("heads", "tails")
n_tries <- 0

repeat {
  # flip coin 10x
  flip <- sample(coin, 10, replace = TRUE)
  
  # add to the number of tries
  n_tries <- n_tries + 1
  
  # if current flip contains all heads or tails break
  if(length(unique(flip)) == 1) {
    print(n_tries)
    break
  }
}
```

## Your Turn!

An elementary example of a random walk is the random walk on the integer number 
line, Z, which starts at 0 and at each step moves +1 or −1 with equal probability.

See if you can do a random walk starting at 0, with each step either adding or subtracting 1. Have your random walk stop if it exceeds 100 or if the number of 
steps taken exceeds 10,000.

Fill in the this incomplete code chunk:

```{r slide-18}
value <- 0
step <- 0
repeat {
  # randomly add or subtract 1
  
  # count step
  
  # break once our walk exceeds 100
  if(some_condition) {
    print(step)
    break
  }
}
```

