---
title: "Scoped dplyr verbs"
output: html_notebook
---

# Module set-up

```{r slide-2}
## ----package-requirements------------------------------------------------
library(nycflights13)  # data
library(dplyr)         # data transformation

## ----data-requirements---------------------------------------------------
flights
```

# dplyr review

* `filter`: pick observations based on values
* `select`: pick variables
* `summarize`: compute statistical summaries
* `group_by`: perform operations at different levels of your data
* `arrange`: reorder data
* `mutate`: create new variables

```{r slide-6}
## ----filter-observations-------------------------------------------------
# filter flights with arrival delay > 60 minutes
filter(flights, arr_delay > 60)
```


```{r slide-7}
## ----select-variables----------------------------------------------------
# select carrier, year, month, day, and all variables containing "delay"
select(flights, carrier, year, month, day, contains("delay"))
```


```{r slide-8}
## ----summary-statistics--------------------------------------------------
# compute median departure delay across all fights
summarize(flights, median_delay = median(dep_delay, na.rm = TRUE))
```


```{r slide-9}
## ----grouped-summary-statistics------------------------------------------
# compute median departure delay for each carrier
flights %>%
  group_by(carrier) %>%
  summarize(median_delay = median(dep_delay, na.rm = TRUE))
```

The `%>%` operator takes the output of the left-hand-side function and passes it 
to the right-hand-side function as the first argument.

```{r slide-10}
#--------------TRADITIONAL APPROACH -------------------------------------------------------
# Step 1: group data
grouped_flights <- group_by(flights, carrier)

# Step 2: compute median departure delay for each carrier
summarize(grouped_flights, median_delay = median(dep_delay, na.rm = TRUE))


#--------------NESTED APPROACH ------------------------------------------------------------
summarize(group_by(flights, carrier), median_delay = median(dep_delay, na.rm = TRUE))


#--------------PIPE OPERATOR APPROACH -----------------------------------------------------
flights %>%
  group_by(carrier) %>%
  summarize(median_delay = median(dep_delay, na.rm = TRUE))
```


```{r slide-11}
## ----arrange-data-based-on-variable--------------------------------------
# rank order carriers with greatest delays on Jan 31
flights %>%
  filter(month == 1, day == 31) %>%
  group_by(carrier) %>%
  summarize(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(avg_delay)) #<<
```


```{r slide-12}
## ----create-new-variable-with-mutate-------------------------------------
# rank order carriers with greatest difference in delays
flights %>%
  filter(month == 1, day == 31) %>%
  group_by(carrier) %>%
  summarize(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
    ) %>%
  mutate(avg_diff = avg_arr_delay - avg_dep_delay) %>% #<<
  arrange(desc(avg_diff ))
```

## Your Turn!

1. Which top 5 destination airports (`dest`) have the largest median arrival 
   delays (`arr_delay`)?
   
2. Which carriers (`carrier`) have the fastest average air speed where 
   `air_speed = distance / air_time`

```{r slide-13}
## ----your-turn-1a--------------------------------------------------------
# 1 
flights %>%
 group_by(____) %>%
 summarize(____) %>% 
 arrange(____)

## ----your-turn-1b--------------------------------------------------------
# 2
flights %>%
 mutate(____) %>%
 group_by(____) %>%
 summarize(____) %>%
  arrange(____)
```


# Scoped dplyr verbs

Performing manipulations on **multiple"** columns can be tedious. Say we wanted 
to compute several departure and arrival time stats for each month:

```{r slide-15}
## ----multiple-summary-statistics-----------------------------------------
# compute several departure and arrival time stats for each month
flights %>%
  group_by(month) %>%
  summarize(
    dep_actual = mean(dep_time, na.rm = TRUE),
    dep_sched  = mean(sched_dep_time, na.rm = TRUE),
    dep_delay  = mean(dep_delay, na.rm = TRUE),
    arr_actual = mean(arr_time, na.rm = TRUE),
    arr_sched  = mean(sched_arr_time, na.rm = TRUE),
    arr_delay  = mean(arr_delay, na.rm = TRUE),
  )
```

Many __dplyr__ functions have cousins with the following suffixes:

- `*_all()`: execute function(s) on all variables
- `*_if()`: execute function(s) on variables that meet a certain condition
- `*_at()`: execute function(s) for pre-specified variables

## Transform all variables with `*all()`

```{r slide-18}
## ----summarize-all-example-1---------------------------------------------
flights %>%
  group_by(month) %>%
  summarize_all(mean, na.rm = TRUE)
```


```{r slide-19}
## ----mutate-all-example-1 -----------------------------------------------
flights %>%
  select(-c(carrier, tailnum, origin, dest, time_hour)) %>%
  mutate_all(scale) 
```

What if we want to write our own custom function on the fly?

```{r slide-20}
## ------------------------------------------------------------------------
flights %>%
  select(-c(carrier, tailnum, origin, dest, time_hour)) %>%
  mutate_all(function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)) #<<

## ------------------------------------------------------------------------
flights %>%
  select(-c(carrier, tailnum, origin, dest, time_hour)) %>%
  mutate_all(.funs = funs((. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE))) #<<
```

## Your Turn!

Using the `flights` data:

1. convert month and day variables to type character
2. group by month and day
3. select all variables containing "time" and "delay"
4. compute the mean of all "time" and "delay" variables

```{r slide-21}
## ---Your-Turn -----------------------------------------------------------
flights %>%
  mutate(____) %>%
  group_by(____) %>%
  select(____) %>%
  summarize_all(____)
```

But what if we only want to transform .red[some] but not all variables?

```{r slide-22}
## ----mutate-all----------------------------------------------------------
# my own standarizing function
standardize <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

# works when performed on only numeric variables
flights %>%
  select(-c(carrier, tailnum, origin, dest, time_hour)) %>%
  mutate_all(standardize)
```

## Transform some variables with `*_if()`

```{r slide-24}
## ----mutate-if-----------------------------------------------------------
# standardize only numeric variables but keep ALL variables
flights %>%
  mutate_if(is.numeric, standardize) 
```

## Your Turn!

Using the `flights` data:

1. convert year, month, and day variables to type character
2. group by month
3. compute the mean of all numeric variables

```{r slide-25}
## ---Your-turn-------------------------------------------------------------
flights %>%
 mutate(____) %>%
 group_by(____) %>%
 summarize_if(____)
```


## Transform specific variables with `*_at()`

```{r slide-28}
## ----unscoped-approach---------------------------------------------------
flights %>%
  mutate(
    year  = as.character(year),  # can we   #<<
    month = as.character(month), # simplify #<<
    day   = as.character(day)    # this?    #<<
  ) %>%
  group_by(month) %>%
  summarize_if(is.numeric, mean, na.rm = TRUE)

## ----scoped-at-approach--------------------------------------------------
flights %>%
  mutate_at(vars(year, month, day), as.character) %>% #<<
  group_by(month) %>%
  summarize_if(is.numeric, mean, na.rm = TRUE)
```

Any `select()` helper function can be used in vars()!

```{r slide-30}
## ----summarize-at-with-helpers-------------------------------------------
flights %>%
  group_by(dest) %>%
  summarize_at(vars(contains("delay"), ends_with("time")), median, na.rm = TRUE)
```

We can also include multiple functions to be executed with the `funs()` wrapper.

```{r slide-31}
## ----summarize-at-with-multiple-functions---------------------------------
flights %>%
  group_by(dest) %>%
  summarize_at(vars(contains("delay")), funs(mean, sd), na.rm = TRUE)
```

## Filtering rows that meet certain conditions

- The `all_vars()` function can be used to filter rows where all variables meet 
  the same logical condition.
  
- The `any_vars()` function can be used to filter rows where at least one variable 
  meets the logical condition.

```{r slide-33}
## ----filter-at-with-all-vars---------------------------------------------
# This will return rows where all variables containing "delay" are NA
flights %>% 
  filter_at(vars(contains("delay")), all_vars(is.na(.)))

## ----filter-at-with-any-vars---------------------------------------------
# This will return rows where any variable containing "delay" is NA
flights %>% 
  filter_at(vars(contains("delay")), any_vars(is.na(.)))
```

## `group_by_*()` and its helpers

```{r slide-34}
## ------------------------------------------------------------------------
flights %>%
  mutate( 
    carrier = as.factor(carrier),
    month   = as.factor(month)
  ) %>%
  group_by(carrier, month) %>%
  summarize_at(vars(contains("delay")), median, na.rm = TRUE)

## ------------------------------------------------------------------------
flights %>%
  group_by_at(vars(carrier, month), as.factor) %>%
  summarize_at(vars(contains("delay")), median, na.rm = TRUE)
```


## Your Turn!

Fill in the blanks and select the right `filter_*()` to filter for those flights where either departure delay (`dep_delay`) or arrival delay (`arr_delay`) exceeded the 99th percentile (hint: `quantile(x, .99)` provides the 99th percentile for variable `x`).

```{r slide-35}
## ---- Your-turn ---------------------------------------------------------
flights %>%
  filter_xxx(vars(contains("_____")), any_vars(___ > quantile(___, .99, na.rm = TRUE)))
```


