---
title: "Conditional Statements"
output: html_notebook
---

# Prerequisites

```{r setup}
library(dplyr)
```


# `if` statement

```{r slide-6}
# create x
set.seed(10)
x <- runif(1)

# check if x meets some condition
if(x > .5) {
  paste("x equals", round(x, 2), "which is greater than 0.5")
}
```

## Dealing with multiple inputs

```{r slide-9}
x <- c(8, 3, -2, 5)

# check if any x elements are less than 0
any(x < 0)

# if statement using any()
if(any(x < 0)) {
  print("x contains a negative number")
}

# check if all x elements are less than 0
all(x < 0)

# if statement using all()
if(all(x < 0)) {
  print("x contains a negative number")
}
```


## Most common uses of `if()` statements

You typically use `if()` statements in functions and looping statements. We'll see more examples and alternatives of these uses in later modules.

```{r slide-10a}
# example `if` statement in function
mean <- function(x) {
  if(!is.numeric(x)) {
    stop("x must be numeric", call. = FALSE)
  }
  sum(x) / length(x)
}
mean(letters)
```

```{r slide-10b}
# example `if` statement in looping statement
# flip a coin until we get 10 heads in a row
results <- NULL
repeat{
  # flip coin & add to results vector
  results <- c(results, sample(c("heads", "tails"), 1))
  # repeat until you have at least 10 flips
  if(length(results) < 10) next()
  # if last 10 flips are all the same
  # report number of flips and break
  if(length(unique(tail(results, 10))) == 1) {
    cat("Number of flips:", length(results))
    break()
  }
}
```


# `ifelse` statement'

1. evaluates if the test expression is TRUE
2. if TRUE, execute statement 1
3. if FALSE, exectute statement 2

```{r slide-12}
# create x
set.seed(10)
x <- runif(1)

# check if x meets some condition
if(x > .7) {
  paste("x equals", round(x, 2), "which is greater than 0.7")
} else {
  paste("x equals", round(x, 2), "which is less than 0.7")
} 
```

We can continue to expand an if...else statement:

```{r}
# testing for more than two conditions
if(x < 3) {
  print("x has low tolerance")
} else if(x == 3 & x < 8) {
  print("x has moderate tolerance")
} else {
  print("x has high tolerance")
}
```

## Your Turn!

Fill in the following code chunk so that:

- if month has value 1-9 the file name printed out will be "data/month-0X.csv"
- if month has value 10-12 the file name printed out will be "data/month-1X.csv"
- if month is an invalid month number (not 1-12), the result printed out is "Invalid month"
test it out for when month equals 6, 10, & 13

```{r slide-16}
month <- 4
if(month _____) {
  paste0("data/", "Month-0", month, ".csv")
} else if(month _____) {
  paste0("data/", "Month-", month, ".csv")
} else {
  print("_____")
}
```


## Vectorized `ifelse`

We can vectorize an if...else statement with:

```{r slide-19a}
# base R approach
x <- c(runif(5), NA)
ifelse(x > .5, "greater than", "less than")
```

```{r slide-19b}
# dplyr R approach
dplyr::if_else(x > .5, "greater than", "less than", missing = "something else")
```

## Vectorize multiple conditional statements

Best way to vectorize multiple conditions (i.e. if...else...if...else]) is with `dplyr::case_when()`

```{r slide-20}
(x <- c(runif(10), NA, Inf))

dplyr::case_when(
  x < .3   ~ "low",
  x < .7   ~ "medium",
  x <= .10 ~ "high",
  is.na(x) ~ "missing",
  TRUE     ~ "out of bounds"
)
```

## Your Turn!

Re-write the below code using a vectorized approach. Test the results with `month <- 1:13`.

```{r slide-21}
month <- 1:13
if(month %in% 1:9) {
  paste0("data/", "Month-0", month, ".csv")
} else if(month %in% 10:12) {
  paste0("data/", "Month-", month, ".csv")
} else {
  print("Invalid month")
}
```


# Conditional statements within dplyr

- So how can we leverage these skills when performing exploratory data analysis?
- Most common is to use use `ifelse()`, `if_else()`, and `case_when()` within `dplyr::mutate()`

```{r slide-22a}
mpg <- ggplot2::mpg

# if_else within mutate
mpg %>%
  select(manufacturer, model, year, cty) %>%
  mutate(cty_rating = if_else(cty > median(cty), "Above", "Below"))
```

```{r slide-22b}
# case_when within mutate
mpg %>%
  select(manufacturer, model, year, cty) %>%
  mutate(cty_rating = case_when(
    cty < quantile(cty, .33) ~ "Low",
    cty < quantile(cty, .77) ~ "Medium",
    TRUE                     ~ "High"
  ))
```

## Your Turn!

Fill in the blanks below to assign each flight to a severity rating of 1, 2, 3, or 4 based on the arrival delay (`arr_delay`) variable:

- severity = 1: if `arr_delay` < 25th percentile
- severity = 2: if `arr_delay` < 50th percentile
- severity = 3: if `arr_delay` < 75th percentile
- severity = 4: if `arr_delay` >= 75th percentile

_Hint: use `quantile(x, perc_value)`_

```{r slide-23}
# fill in the blanks
nycflights13::flights %>%
  filter(arr_delay > 0) %>%
  select(carrier, tailnum, arr_delay) %>%
  mutate(severity = case_when(
    ______ ~ 1,
    ______ ~ 2,
    ______ ~ 3,
    ______ ~ 4
  ))
```

