---
title: "Iteration with Functional Programming"
output: html_notebook
---


The __purr__ package provides functional programming tools that:

1. for each element of x
2. apply function f and
3. provide consistent output

# Prerequisites

```{r slide-9}
library(tidyverse)    # includes purrr and other pkgs we'll use
library(repurrrsive)  # provides different data sets
```

## Your Turn!

Run the following examples in your console. Don't worry about understanding 
everything, but identify

1. what is the input of each map function,
2. what is the output of each map function, and
3. talk to your neighbor about what you think each function is doing.

```{r slide-14}
# 1.
iris %>% map(n_distinct)

# 2.
input <- c(1, 5, 10)
input %>% map(runif, min = 1, max = 2)

# 3.
input %>% map(~ sample(1:10, size = .x, replace = TRUE))
```


# The function call

We can make function calls in different ways:

```{r slide-16-19}
# function with default parameters, by default
# .x will go in as the first argument
map(input, runif)

# same as above but supplying additional parameters
map(input, runif, min = 1, max = 2)

# anonymous function defined using formula shortcut
# handy when .x is not first argument
map(input, ~ sample(1:10, size = .x, replace = TRUE))

# traditional approach to apply anonymous function
# TOO VERBOSE!
map(input, function(x) sample(1:10, size = .x, replace = TRUE))
```


## Your Turn!

Remember how `list.files("data")` shows all the .csv files? See if you can:

1. combine `list.files("data")` and `paste0("data/", ...)` to create the following inputs:

```
##  [1] "data/Month-01.csv" "data/Month-02.csv" "data/Month-03.csv"
##  [4] "data/Month-04.csv" "data/Month-05.csv" "data/Month-06.csv"
##  [7] "data/Month-07.csv" "data/Month-08.csv" "data/Month-09.csv"
## [10] "data/Month-10.csv" "data/Month-11.csv"
```

2. now use this input vector, `map()`, and `read_csv()` to import all the data 
files into a list

```{r slide-21}
list.files("data") %>% 
  paste0("data/", .) %>%
  map(______)
```


# Different `map()` functions for different outputs

As previously mentioned, `map()` returns a list, which makes it the most general 
of the “map” family because you can put anything in a list.

It is inconvenient to return a list when a simpler data structure would do

Alternative map() variants include:
   - `map_dbl()` returns a double vector
   - `map_lgl()` returns a logical vector
   - `map_int()` returns a integer vector
   - `map_chr()` returns a character vector
   - `map_df()` returns a data frame

- Its best to use the most restrictive function for the output
- Informative errors will let you know when a problem arises with your output
- You can always use `map()` to help debug the output error

```{r slide-23}
# output is double
mtcars %>% map_dbl(sd)
 
# output is integer
iris %>% map_int(n_distinct)
 
# output is logical
iris %>% map_lgl(is.numeric)
 
# output is character
iris %>% map_chr(is.character)

# output is data frame
mtcars %>% map_df(scale)
```

## Your Turn!

With the built-in airquality data set, use the most appropriate map functions to 
answer these three questions:

1. how many n_distinct values are in each column?
2. are there any missing values in this data set?
3. what is the standard deviation for each variable?

```{r slide-24}

```


# Road `map()` Strategy

1. Do it for one element
2. Turn it into a recipe
3. Use `map_*()` to do it to all elements

## How many films has each character been in?

- the sw_people data set contains all the characters from the Star Wars chronicles
- What if we want to find how many $films each character has been in?

```{r slide-27}
glimpse(sw_people)
```

## #1 Do it for one element

```{r slide-28}
# do it for the first element
x1 <- sw_people[[1]]
length(x1$films)

# do it for the second element
x2 <- sw_people[[2]]
length(x2$films)

# do it for the third element
x3 <- sw_people[[3]]
length(x3$films)
```

## #2 Turn it into a recipe

- `~`: anonymous function shortcut
- `.x`: one element of our vector

```{r slide-33}
sw_people %>% map_int(~ length(.x$films))
```

## Your Turn!

Use the:

1. do it for one,
2. create recipe,
3. do it for all

strategy to:

1. Extract the name of each character
2. Identify if the character is male
3. Identify the proportion of characters that are male

```{r slide-34}
# use sw_people to answer the above questions
sw_people
```


# Some advanced features to whet your appetite for more!

## Renaming elements

Pro tip: purrr has a `set_names()` function that makes renaming elements easy

```{r slide-39}
sw_people %>% 
  map_chr(~ .x$name) %>%          # get names of each character
  set_names(sw_people, .) %>%     # set list names
  map_int(~ length(.x$films)) %>% # get number of films
  sort(decreasing = TRUE) 
```

## Combining multiple functionals is powerful. 

Say we wanted to compare the approval/disapproval ratio for Transaction_Status 
across all our Month-XX.csv files.

We can accomplish this with a sequence of functionals as follows:

```{r slide-40a}
list.files("data") %>% 
  paste0("data/", .) %>%
  map(read_csv) %>%
  set_names(., nm = list.files("data")) %>%
  map(~ .x$Transaction_Status == "Approved") %>%
  map_dbl(mean)
```

Or with some initial functionals followed by __dplyr__:

```{r slide-40b}
list.files("data") %>% 
  paste0("data/", .) %>%
  map_dfr(read_csv) %>%  # row bind list to data frame
  group_by(Month) %>%
  summarize(Approval_ration = mean(Transaction_Status == "Approved"))
```

## `map2()` for two input elements

Say we have a procedure based on two inputs. For example, we want to make 
predictions for 10 countries with their respective linear models.

```{r slide-41}
# input 1: data
glimpse(gap_split_small)

# input 2: models
(gap_mods <- gap_split_small %>%
  map(~ lm(
    formula = lifeExp ~ year + pop + gdpPercap,
    data = .x
    ))
)
```

```{r slide-42a}
# 1. do for one
predict(gap_mods[[1]], gap_split_small[[1]])
```

```{r slide-42b, eval=FALSE}
# 2. identify recipe
predict(________, ________)
```

```{r slide-42c}
# 3. do for all
map2(gap_mods, gap_split_small, predict) %>% glimpse(give.attr = FALSE)
```

## `pmap()` for n input elements

And if we have a procedure with $>2$ elements:

Say we want to iterate over each of the following hyperparameters for a random forest model:

```{r slide-43}
# list of hyper parameter inputs
hyper_params <- list(
  mtry = c(20, 25, 30),
  node_size = c(3, 6, 9),
  sample_size = c(.50, .632, .80)
)
```

```{r slide-44a}
# 1. do it for one
ranger::ranger(
  formula         = Sale_Price ~ ., 
  data            = AmesHousing::make_ames(), 
  num.trees       = 500,
  mtry            = hyper_params$mtry[1],
  min.node.size   = hyper_params$node_size[1],
  sample.fraction = hyper_params$sample_size[1],
  seed            = 123
  )$prediction.error
```

```{r slide-44b, eval=FALSE}
# 2. identify recipe
ranger::ranger(
  formula         = Sale_Price ~ ., 
  data            = AmesHousing::make_ames(), 
  num.trees       = 500,
  mtry            = ________,
  min.node.size   = ________,
  sample.fraction = ________,
  seed            = 123
  )$prediction.error
```

```{r slide44-c}
# 3. do it for all
pmap_dbl(
  hyper_params, 
  ~ ranger::ranger(
    formula         = Sale_Price ~ ., 
    data            = AmesHousing::make_ames(), 
    num.trees       = 500,
    mtry            = ..1,
    min.node.size   = ..2,
    sample.fraction = ..3,
    seed            = 123
    )$prediction.error
  )
```





