---
title: "Iterating with Loops"
output:
  xaringan::moon_reader:
    css: ["scrollable.css", "mtheme_max.css", "fonts_mtheme_max.css"]
    self_contained: false
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false   
---

```{r setup, include=FALSE, cache=FALSE}
# Set global R options
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)

# Set global knitr chunk options
knitr::opts_chunk$set(
  fig.align = "center", 
  cache = TRUE,
  error = FALSE,
  message = FALSE, 
  warning = FALSE, 
  collapse = TRUE 
)

# set ggplot theme in case I do any plotting
library(ggplot2)
thm <- theme_bw()
theme_set(thm)
```

class: clear, center, middle

background-image: url(http://static1.squarespace.com/static/57d161dc20099eb4bb0ccc05/t/57d69e61414fb51a88bd4798/1523877244472/)
background-size: contain

---
# Intro

.font120[
Often, we need to execute repetitive code statements a particular number of times. Or, we may even need to execute code for an undetermined number of times until a certain condition no longer holds.
]

--

* .font130[`for`] loops: execute code for a prescribed number of times (Rule of `r emo::ji("+1")`, 3x).

* .font130[`while`] & .font130[`repeat`] loops: repeat code while condition holds

* .font130[`next`] & .font130[`break`]: additional clauses to control flow

--

<br>

.center.bold[There are additional iterating functions which we'll learn about in the functional programming module]

---
# `for` loop: syntax

.pull-left[
.font120[
* For each element in sequence (1-100) perform some defined task
]
]

.pull-right[

```{r, eval=FALSE}
for(i in 1:100) {
  <do stuff here with i>
}
```

]

---
# `for` loop: syntax

.pull-left[
* For each element in sequence (1-100) perform some defined task

* .red[What will this .font120[`for`] loop do?]
]

.pull-right[

```{r, eval=FALSE}
for(i in 1:100) {
  <do stuff here with i>
}
```

```{r, eval=FALSE}
for (i in 2010:2017) {
  print(paste("The year is", i))
}
```
]

---
# `for` loop: syntax

.pull-left[
* For each element in sequence (1-100) perform some defined task

* .red[What will this .font120[`for`] loop do?]
]

.pull-right[

```{r, eval=FALSE}
for(i in 1:100) {
  <do stuff here with i>
}
```

```{r}
for (i in 2010:2017) {
  print(paste("The year is", i))
}
```
]

---
# `for` loop: sequence input

.pull-left[
* Often, the length of the sequence is determined by some pre-defined variable

* For example, say we have a variable .font120.blue[`x`] and we want to iterate over each element and do something

* We can do this two ways

]

.pull-right[

```{r, eval=FALSE}
x <- 2010:2017 #<<

# option 1
for(i in 1:length(x)) {
  <do stuff here with i>
}

# option 2
for(i in seq_along(x)) {
  <do stuff here with i>
}
```

]

<br>
.center.bold.red[Talk to your neighbor. Do you prefer one over the other? `r anicon::cia("https://emojis.slackmojis.com/emojis/images/1504806416/2885/thinking-face-rotating.gif?1504806416", animate = FALSE, size = 2)`
]

---
# `for` loop: sequence input

.pull-left[
* Often, the length of the sequence is determined by some pre-defined variable

* For example, say we have a variable .font120.blue[`x`] and we want to iterate over each element and do something

* We can do this two ways

* But `seq_along(x)` is safer for edge cases where the sequence has length zero

]

.pull-right[

```{r}
x <- c()

# option 1: executes although it shouldn't
for(i in 1:length(x)) {
  print(i)
}

# option 2: does not execute
for(i in seq_along(x)) {
  print(i)
}
```

]

<br>

.center.bold.red[Beware iterating over `1:length(x)`. This will fail in unhelpful ways if `x` has length 0]

---
# `for` loop: generating output

.pull-left[

* Often we are generating output from the `for` loop that we want to use later

* We can do this by either .bold.red[growing] or .bold.green[filling] the output

* Pre-allocating the output and .bold.green[filling] in the results is more efficient

]

.pull-right[

.code70[
```{r}
# Let's create 100 different vectors w/10,000 obs
means <- seq(0, 100, by = 1)

# Growing the output
out <- list()
system.time({
  for(i in seq_along(means)) {
    # generate data
    gen_data <- rnorm(10000, means[[i]])
    # grow list output
    out <- c(out, gen_data)
  }
})


# Pre-allocating and filling
out <- vector("list", length(means))
system.time({
  for(i in seq_along(means)) {
    # generate & fill output
    out[[i]] <- rnorm(10000, means[[i]])
  }
})
```
]
]

---
class: yourturn
# Your Turn: Challenge

* We can see all data sets that we have in the "data" folder with `list.files()`

.code70[
```{r}
list.files("data")
```
]

.pull-left[

* Say we wanted to import one of these files into R

```{r}
# here's a single file
(first_df <- list.files("data")[1])

# create path and import this single file
df <- readr::read_csv(paste0("data/", first_df))

# create a new name for file
(new_name <- stringr::str_sub(first_df, end = -5))

# dynamically rename file
assign(new_name, df)
```

]

--

.pull-right[

* Can you incorporate these procedures into a `for` loop to import all the data files?

```{r, eval=FALSE}
for(i in _____) {
  # 1: import data
  df <- readr::read_csv(_____)
  
  # 2: remove ".csv" of name
  new_name <- _____
  
  # 3: dynamically rename file
  assign(_____, _____)
}
```


]


---
class: yourturn
# Your Turn: Solution

.pull-left[

1. Import ith data file

2. Make a new name without ".csv" (i.e. "Month-01")

3. Dynamically rename file with `assign()`

4. Remove unnecessary duplicate

]

.pull-right[

```{r}
for(i in list.files("data")) {
  # 1: import data
  df <- readr::read_csv(paste0("data/", i))
  
  # 2: remove ".csv" of name
  new_name <- stringr::str_sub(i, end = -5)
  
  # 3: dynamically rename file
  assign(new_name, df)
  
  # 4: remove unnecessary object
  rm(df)
}
```

]

---
# Controlling sequences

.pull-left[

There are two ways to control the progression of a loop:

.font120[

* `next`: terminates the current iteration and advances to the next.

]
]

.pull-right[

```{r}
x <- c(1, 2, NA, 3)

# use next to skip an iteration
for (i in x) {
  if (is.na(i)) next    # if NA don't execute #<<

  print(i)
}
```

]

---
# Controlling sequences

.pull-left[

There are two ways to control the progression of a loop:

* `next`: terminates the current iteration and advances to the next.

* `break`: exits the entire `for` loop.

]

.pull-right[

```{r}
x <- c(1, 2, NA, 3)

# use next to skip an iteration
for (i in x) {
  if (is.na(i)) next

  print(i)
}
```


```{r}
# use break to stop
for (i in x) {
  if (is.na(i)) break     # if NA stop! #<<

  print(i)
}
```

]

--

.center.red[_both are used in conjunction with `if` statements_]

---
class: yourturn
# Your Turn: Challenge!

.pull-left[

The following code identifies the month of the data set:

```{r}
# data files
(data_files <- list.files("data"))

# extract month number
as.numeric(stringr::str_extract(data_files, "\\d+"))
```

]

--

.pull-right[

Modify the following `for` loop with a `next` or `break` statement to:

1. only import `Month-01` through `Month-07`
2. only import `Month-08` through `Month-10`

```{r, eval=FALSE}
# Modify this code chunk with you next/break statement
for(i in list.files("data")) {
  # steps to import each data set
  df <- readr::read_csv(paste0("data/", i))
  new_name <- stringr::str_sub(i, end = -5)
  assign(new_name, df)
  rm(df)
}
```


]

---
class: yourturn
# Your Turn: Solution!

.pull-left[

.center[\#1: only import `Month-01` through `Month-07`]

```{r, echo=FALSE, cache=FALSE}
rm(list = ls())
```


```{r, cache=FALSE}
for(i in list.files("data")) {
  
  # break if month is greater than 7; although #<<
  # this requires knowledge of month order  #<<
  month <- as.numeric(stringr::str_extract(i, "\\d+")) #<<
  if(month > 7) break #<<
  
  # steps to import each data set
  df <- readr::read_csv(paste0("data/", i))
  new_name <- stringr::str_sub(i, end = -5)
  assign(new_name, df)
  rm(df)

}

# we've only imported months 1-7
ls(pattern = "Month")
```

]

.pull-right[

.center[\#2: only import `Month-08` through `Month-10`]

```{r, echo=FALSE, cache=FALSE}
rm(list = ls())
```

```{r, cache=FALSE}
for(i in list.files("data")) {
  
  # skip importing file if month value is not 8-10 #<<
  month <- as.numeric(stringr::str_extract(i, "\\d+")) #<<
  if(month < 8 | month > 10) next #<<
  
  # steps to import each data set
  df <- readr::read_csv(paste0("data/", i))
  new_name <- stringr::str_sub(i, end = -5)
  assign(new_name, df)
  rm(df)
  
}

# we've only imported months 1-7
ls(pattern = "Month")
```

]

---
# `while` loop


---
# `repeat` loop


---
# Summary


---
# Questions











