---
title: "Iterating with Loops"
output:
  xaringan::moon_reader:
    css: ["scrollable.css", "mtheme_max.css", "fonts_mtheme_max.css"]
    self_contained: false
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false   
---

```{r setup, include=FALSE, cache=FALSE}
# Set global R options
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)

# Set global knitr chunk options
knitr::opts_chunk$set(
  fig.align = "center", 
  cache = TRUE,
  error = FALSE,
  message = FALSE, 
  warning = FALSE, 
  collapse = TRUE 
)

# set ggplot theme in case I do any plotting
library(ggplot2)
thm <- theme_bw()
theme_set(thm)
```

class: clear, center, middle

background-image: url(http://static1.squarespace.com/static/57d161dc20099eb4bb0ccc05/t/57d69e61414fb51a88bd4798/1523877244472/)
background-size: contain

---
# Intro

.font120[
Often, we need to execute repetitive code statements a particular number of times. Or, we may even need to execute code for an undetermined number of times until a certain condition no longer holds.
]

--

* .font130[`for`] loops: execute code for a prescribed number of times (Rule of `r emo::ji("+1")`, 3x).

* .font130[`while`] & .font130[`repeat`] loops: repeat code while condition holds

* .font130[`next`] & .font130[`break`]: additional clauses to control flow

--

<br>

.center.bold[There are additional iterating functions which we'll learn about in the functional programming module]

---
# `for` loop: syntax

.pull-left[
.font120[
* For each element in sequence (1-100) perform some defined task
]
]

.pull-right[

```{r, eval=FALSE}
for(i in 1:100) {
  <do stuff here with i>
}
```

]

---
# `for` loop: syntax

.pull-left[
* For each element in sequence (1-100) perform some defined task

* .red[What will this .font120[`for`] loop do?]
]

.pull-right[

```{r, eval=FALSE}
for(i in 1:100) {
  <do stuff here with i>
}
```

```{r, eval=FALSE}
for (i in 2010:2017) {
  print(paste("The year is", i))
}
```
]

---
# `for` loop: syntax

.pull-left[
* For each element in sequence (1-100) perform some defined task

* .red[What will this .font120[`for`] loop do?]
]

.pull-right[

```{r, eval=FALSE}
for(i in 1:100) {
  <do stuff here with i>
}
```

```{r}
for (i in 2010:2017) {
  print(paste("The year is", i))
}
```
]

---
# `for` loop: sequence input

.pull-left[
* Often, the length of the sequence is determined by some pre-defined variable

* For example, say we have a variable .font120.blue[`x`] and we want to iterate over each element and do something

* We can do this two ways

]

.pull-right[

```{r, eval=FALSE}
x <- 2010:2017 #<<

# option 1
for(i in 1:length(x)) {
  <do stuff here with i>
}

# option 2
for(i in seq_along(x)) {
  <do stuff here with i>
}
```

]

<br>
.center.bold.red[Talk to your neighbor. Do you prefer one over the other? `r anicon::cia("https://emojis.slackmojis.com/emojis/images/1504806416/2885/thinking-face-rotating.gif?1504806416", animate = FALSE, size = 2)`
]

---
# `for` loop: sequence input

.pull-left[
* Often, the length of the sequence is determined by some pre-defined variable

* For example, say we have a variable .font120.blue[`x`] and we want to iterate over each element and do something

* We can do this two ways

* But `seq_along(x)` is safer for edge cases where the sequence has length zero

]

.pull-right[

```{r}
x <- c()

# option 1: executes although it shouldn't
for(i in 1:length(x)) {
  print(i)
}

# option 2: does not execute
for(i in seq_along(x)) {
  print(i)
}
```

]

<br>

.center.bold.red[Beware iterating over `1:length(x)`. This will fail in unhelpful ways if `x` has length 0]

---
# `for` loop: generating output

.pull-left[

* Often we are generating output from the `for` loop that we want to use later

* We can do this by either .bold.red[growing] or .bold.green[filling] the output

* Pre-allocating the output and .bold.green[filling] in the results is more efficient

]

.pull-right[

.code70[
```{r}
# Let's create 100 different vectors w/10,000 obs
means <- seq(0, 100, by = 1)

# Growing the output
out <- list()
system.time({
  for(i in seq_along(means)) {
    # generate data
    gen_data <- rnorm(10000, means[[i]])
    # grow list output
    out <- c(out, gen_data)
  }
})


# Pre-allocating and filling
out <- vector("list", length(means))
system.time({
  for(i in seq_along(means)) {
    # generate & fill output
    out[[i]] <- rnorm(10000, means[[i]])
  }
})
```
]
]

---
class: yourturn
# Your Turn: Challenge

* We can see all data sets that we have in the "data" folder with `list.files()`

.code70[
```{r}
list.files("data")
```
]

.pull-left[

* Say we wanted to import one of these files into R

```{r}
# here's a single file
(first_df <- list.files("data")[1])

# create path and import this single file
df <- readr::read_csv(paste0("data/", first_df))

# create a new name for file
(new_name <- stringr::str_sub(first_df, end = -5))

# dynamically rename file
assign(new_name, df)
```

]

--

.pull-right[

* Can you incorporate these procedures into a `for` loop to import all the data files?

```{r, eval=FALSE}
for(i in _____) {
  # 1: import data
  df <- readr::read_csv(_____)
  
  # 2: remove ".csv" of name
  new_name <- _____
  
  # 3: dynamically rename file
  assign(_____, _____)
}
```


]


---
class: yourturn
# Your Turn: Solution

.pull-left[

1. Import ith data file

2. Make a new name without ".csv" (i.e. "Month-01")

3. Dynamically rename file with `assign()`

4. Remove unnecessary duplicate

]

.pull-right[

```{r}
for(i in list.files("data")) {
  # 1: import data
  df <- readr::read_csv(paste0("data/", i))
  
  # 2: remove ".csv" of name
  new_name <- stringr::str_sub(i, end = -5)
  
  # 3: dynamically rename file
  assign(new_name, df)
  
  # 4: remove unnecessary object
  rm(df)
}
```

]

---
# Controlling sequences

.pull-left[

There are two ways to control the progression of a loop:

.font120[

* `next`: terminates the current iteration and advances to the next.

]
]

.pull-right[

```{r}
x <- c(1, 2, NA, 3)

# use next to skip an iteration
for (i in x) {
  if (is.na(i)) next    # if NA don't execute #<<

  print(i)
}
```

]

---
# Controlling sequences

.pull-left[

There are two ways to control the progression of a loop:

* `next`: terminates the current iteration and advances to the next.

* `break`: exits the entire `for` loop.

]

.pull-right[

```{r}
x <- c(1, 2, NA, 3)

# use next to skip an iteration
for (i in x) {
  if (is.na(i)) next

  print(i)
}
```


```{r}
# use break to stop
for (i in x) {
  if (is.na(i)) break     # if NA stop! #<<

  print(i)
}
```

]

--

.center.red[_both are used in conjunction with `if` statements_]

---
# Warning `r anicon::faa("exclamation-triangle", animate = FALSE, color = "red")`

Note that `break` and `next` only control the progression of the .red.bold[direct] loop environment they are in.

.pull-left[

.center[Break out of inner loop only]

```{r}
x <- 1:3
y <- c("a", "b", NA)

for(i in x) {
  print(i)
  for(j in y) {
    if(is.na(j)) break # break out of inner loop #<<
    print(j)
  }
}
```


]

.pull-right[

.center[Stop entire process.]

```{r, error=TRUE}
x <- 1:3
y <- c("a", "b", NA)

for(i in x) {
  print(i)
  for(j in y) {
    if(is.na(j)) stop("break in inner loop") #<<
    print(j)
  }
}
```

.center.font80.red[You'll learn more about `stop()` in the functions module!]

]

---
class: yourturn
# Your Turn: Challenge!

.pull-left[

The following code identifies the month of the data set:

```{r}
# data files
(data_files <- list.files("data"))

# extract month number
as.numeric(stringr::str_extract(data_files, "\\d+"))
```

]

--

.pull-right[

Modify the following `for` loop with a `next` or `break` statement to:

1. only import `Month-01` through `Month-07`
2. only import `Month-08` through `Month-10`

```{r, eval=FALSE}
# Modify this code chunk with you next/break statement
for(i in list.files("data")) {
  # steps to import each data set
  df <- readr::read_csv(paste0("data/", i))
  new_name <- stringr::str_sub(i, end = -5)
  assign(new_name, df)
  rm(df)
}
```


]

---
class: yourturn
# Your Turn: Solution!

.pull-left[

.center[\#1: only import `Month-01` through `Month-07`]

```{r, echo=FALSE, cache=FALSE}
rm(list = ls())
```


```{r, cache=FALSE}
for(i in list.files("data")) {
  
  # break if month is greater than 7; although #<<
  # this requires knowledge of month order  #<<
  month <- as.numeric(stringr::str_extract(i, "\\d+")) #<<
  if(month > 7) break #<<
  
  # steps to import each data set
  df <- readr::read_csv(paste0("data/", i))
  new_name <- stringr::str_sub(i, end = -5)
  assign(new_name, df)
  rm(df)

}

# we've only imported months 1-7
ls(pattern = "Month")
```

]

.pull-right[

.center[\#2: only import `Month-08` through `Month-10`]

```{r, echo=FALSE, cache=FALSE}
rm(list = ls())
```

```{r, cache=FALSE}
for(i in list.files("data")) {
  
  # skip importing file if month value is not 8-10 #<<
  month <- as.numeric(stringr::str_extract(i, "\\d+")) #<<
  if(month < 8 | month > 10) next #<<
  
  # steps to import each data set
  df <- readr::read_csv(paste0("data/", i))
  new_name <- stringr::str_sub(i, end = -5)
  assign(new_name, df)
  rm(df)
  
}

# we've only imported months 1-7
ls(pattern = "Month")
```

]

---
# Repeating code for undefined number of iterations

Sometimes we need to execute code for an undetermined number of times until a certain condition no longer holds.

Two very similar options:

--

.pull-left[

.center.bold[while loop]

```{r, eval=FALSE}
while(condition) {
  <do stuff>
}
```

<div style="height:15px;font-size:15px;">&nbsp;</div>

* Test condition first

* Then execute code

]

--

.pull-right[

.center.bold[repeat loop]

```{r, eval=FALSE}
repeat {
  <do stuff>
    
  if(condition) break  
}
```

* Execute code first

* Then test condition

]

---
# Repeating code for undefined number of iterations

The probability of flipping 10 coins and getting all heads or tails is $(\frac{1}{2})^{10} = 0.0009765625$ or 1 in 1024 tries. 

Can we beat these odds?

.pull-left[

```{r}
coin <- c("heads", "tails")
n_tries <- 0
flip <- NULL

while(length(unique(flip)) != 1) {
  # flip coin 10x
  flip <- sample(coin, 10, replace = TRUE)
  
  # add to the number of tries
  n_tries <- n_tries + 1
}

n_tries
```


]

--

.pull-right[

```{r}
coin <- c("heads", "tails")
n_tries <- 0

repeat {
  # flip coin 10x
  flip <- sample(coin, 10, replace = TRUE)
  
  # add to the number of tries
  n_tries <- n_tries + 1
  
  # if current flip contains all heads or tails break
  if(length(unique(flip)) == 1) {
    print(n_tries)
    break
  }
}
```


]

---
class: yourturn
# Your Turn: Challenge!

.pull-left[
An elementary example of a random walk is the random walk on the integer number line, $\mathbb {Z}$ , which starts at 0 and at each step moves +1 or −1 with equal probability.

See if you can do a random walk starting at 0, with each step either adding or subtracting 1.  Have you random walk stop if it exceeds 100 or if the number of steps taken exceeds 10,000.
]

--

.pull-right[

Fill in this incomplete code chunk:

```{r, eval=FALSE}
value <- 0
step <- 0

repeat {
  # randomly add or subtract 1

  # count step

  # break once our walk exceeds 100
  if(some_condition) {
    print(step)
    break
  }
}
```

]

---
class: yourturn
# Your Turn: Solution!

.pull-left[

This example reached 100 in 5,666 steps

```{r}
value <- 0
step <- 0

repeat {
  # randomly add or subtract 1
  random_step <- sample(c(-1, 1), 1)
  value <- value + random_step
  
  # count step
  step <- step + 1
  
  # break once our walk exceeds 100
  if(value == 100 | step > 10000) {
    print(step)
    break
  }
}
```

]

--

.pull-right[

What does a random walk look like? This example never reached 100.

```{r, echo=FALSE, fig.height=4}
value <- 0
step <- 0

repeat {
  # count step
  step <- step + 1
  
  # randomly add or subtract 1
  random_step <- sample(c(-1, 1), 1)
  value[step + 1] <- value[step] + random_step
  
  # break once our walk exceeds 100
  if(max(value) >= 100 | step > 10000) {
    break
  }
}

library(ggplot2)
df <- data.frame(
  step = c(0, seq_len(step)),
  value = value
)

ggplot(df, aes(step, value)) +
  geom_line()
```

In fact, it looks a little like the current stock market, eh?  

.center[`r anicon::cia("https://emojis.slackmojis.com/emojis/images/1502933663/2802/homer_scream.jpg?1502933663", animate = FALSE, size = 2)`]

]

---
# Summary


---
# Questions











